The Fremont Frontier

### 2024-01-31 manuscript/cover-letter-response-to-reviews.md

We thank the reviewers for their many insightful comments and criticisms
regarding our paper.

All minor comments about grammatical errors, poor wording, and the like have
been addressed in the text. Major comments from reviewers are listed here with
comments about how we addressed them in the text. Reviewer two had the most
substantial comments and criticisms, so most of this involves responses to them.

### Reviewer #1

1.  Description of Fremont extent is misleading and vague.

    Response: We updated the description of the Fremont extent in the background
    section to make it clearer that there is evidence for Fremont foragers in ID
    and WY and Fremont farmers in NV and CO, rather than leaving it vague
    whether they might have all been farmers.

2.  Description of project area is inaccurate.

    Response: We updated this section to fix inaccurate descriptions. We also
    updated the overview map, adding major roads and some important nearby
    cities to give the reader more landmarks for orientation.

3.  The impossibility of dry farming does not necessarily entail irrigation.

    Response: This is correct, so we have added text to the discussion to
    clarify. In the subsection title, we also replaced 'irrigation' with 'water
    management', and in the text, we change "making irrigation ... necessary" to
    "making irrigation or some other form of water management ... necessary."

### Reviewer #2

1.  Why not high precipitation?

    Response: Our answer to this involves the trade-off with temperature in the
    project area. As elevation increases, precipitation increases but
    temperature decreases. So, the places with the most precipitation are also
    the ones most likely to experience frost and to have shorter growing
    seasons. This means the best place for maize is not necessarily the place
    with the most precipitation. We've updated the discussion, rearranging some
    comments into a new section entitled "Trade-off Between Precipitation and
    Temperature," to emphasize this point.

2.  The limits of rain-fed maize farming should be discussed more explicitly.

    Response: Our analysis of the Fremont maize farming niche is suggestive of
    its limits, but defining the true limits requires actualistic research,
    hence our reluctance to offer specific numbers on this subject. We do,
    however, discuss the results of some actualistic research, particularly the
    work referenced by Bocinsky in his 2014, and compare our results to that.
    This is the basis for our claim that the Fremont likely lived outside of the
    rain-fed maize farming niche, thus requiring some kind of water management
    on their part.

3.  Details of data collection are too vague.

    Response: We updated the Site Data section to more accurately represent how
    we came by these data.

    -   The NSF project alluded to in the previous version of this paper now
        includes its award number along with several citations to research
        coming out of that project.
    -   The "stratified random sample" comment was a bit misleading. That was
        done for a portion of the sites (we sampled a fraction of site forms in
        each county), but not systematically to produce the entire sample, which
        was mostly arrived at in a more opportunistic manner.
    -   To get room counts, we used the list of Fremont villages found in
        Coltrain and Leavitt (2002), and cross-referenced those with the
        original site forms, excavation reports, and Adrien Mooney's 2014 BYU MA
        thesis, which tries to reconstruct some poorly reported excavations. For
        many of these sites, the estimates can vary wildly, so we erred on the
        side of choosing the more conservative estimates.
    -   In spite of these limitations with our data collection scheme, we note
        that the current dataset represents one of the largest Fremont samples
        used for this sort of analysis to date.

4.  What kind of sites are these?

    Response: The reviewer requests additional information about site types, but
    we do not currently have the data to answer that question. Mainly, our data
    collection efforts were focused on identifying as many Fremont related sites
    as possible. In fact, we have done this now for a considerable portion of
    the state of Utah, though efforts are ongoing. We are also trying to go back
    and record as much attribute data as is available and will hopefully be able
    to report on these efforts in the near future.

5.  How do we know that these are farming and not foraging adaptations?

    Response: The reviewer notes that site types are needed to differentiate
    hunting sites from farming sites, and thus to determine whether settlement
    decisions were shaped by constraints on farming vs constraints on foraging.
    This is a really important point that goes right to the heart of our model
    interpretation. If the sites in our analysis are not primarily associated
    with farmers, then we cannot use them to understand the maize farming niche.
    Unfortunately, there's just not a lot of good data to really answer this
    question. The available stable isotope data that exists for the Fremont
    suggests that they were, for the most part, consuming maize at a rate
    comparable to maize farmers in the San Juan Basin, including Chaco Canyon.
    See, for example, Coltrain and Leavitt 2002 and Coltrain 1993. As Coltrain
    and Leavitt point out, the one exception to this general pattern is the
    Fremont around Bear River and the Willard Bay wetlands, but even there
    evidence of a largely hunting based diet is limited to the last few
    centuries in their sequence (around 900 CE their subsistence pattern becomes
    more mixed, and only around 1100 do they fully adopt foraging). This is not
    to say that foraging wasn't important to the Fremont. Just that farming was
    likely the critical constraint on settlement for most Fremont for most of
    their history (Yaworsky 2023; Spangler 2000, 2002).

6.  The Ideal Free Distribution section is overly detailed.

    Response: The reviewer thinks the discussion of the IFD model is overly
    detailed because "the analysis doesn't rank habitat suitability or site
    association with suitability." The reviewer suggests a simpler discussion
    that contextualizes the analysis "within the literature on SDMs." In
    response, we would like to note three things.

    -   First, we agree that the section was overly detailed. In fact, the main
        logic we wanted to convey is already in the intro, so the section is not
        just overly detailed but basically unnecessary, so we have removed it
        entirely.

    -   Second, distressingly, it's precisely the goal of this paper to estimate
        relative differences in habitat suitability across watersheds and
        environmental gradients, but the reviewer's comment suggests that we did
        not do a good job of making this clear. So, we have updated the
        introduction (particularly, the last paragraph) and the discussion to
        make this goal more explicit.

    -   Third, as the reviewer requested, we also explicitly connected this to
        the SDM literature by noting that ecologists will sometimes refer to
        SDMs as "Habitat Suitability Models" (e.g., Rowden et al 2017; Kort et
        al 2020; Bowden et al 2021).

7.  Goldilocks Zone concept is misleading.

    Response: The title of the subsection was changed to "Trade-off Between
    Precipitation and Temperature." The expression is now described explicitly
    as a "metaphor," and it's made clear that the reader should not
    over-interpret it for the reasons listed by the reviewer. We have also
    removed all but a single reference to it.

### Reviewer #4

1.  There are several reference errors and omissions.

    Response: These have been corrected. The more recent references to Fremont
    scholarship offered by the reviewer have also been added.

2.  But are they optimizers?

    Response: This paper is not meant to be a defense of the optimality
    assumption, just an example of its application, so we would prefer not to
    litigate that question here. We simply propose that Fremont farmers located
    themselves in places that yield the highest probabilities of a successful
    harvest. This assumption only requires that the Fremont had an intimate
    knowledge of their local ecology and leveraged this knowledge to make
    intelligent decisions regarding the placement of farming sites. Some of the
    wording in the introduction was changed to emphasize this.

3.  Trade-offs between spatial and temporal resolution?

    Response: We added an entire section to the discussion on this specific
    topic. It has the title "The Data Quality Problem in Fremont Research."
    Hopefully, it helps to assuage the reviewer's concern about a lack of
    temporal resolution in our data, which we fully admit is an issue. While
    having precise chronological control would be preferable, we argue that our
    results are not meaningless because "typically better watersheds," meaning
    watersheds that were warmer and wetter for most of the Fremont sequence,
    "would have attracted maize farmers more often and, over time, would have
    accumulated a larger archaeological maize farming assemblage, too." So, the
    results aren't perfect, but they are at least suggestive of the real drivers
    of Fremont residential choice behavior.

### 2024-01-31 manuscript/manuscript.qmd
all changes made to meet reviewer concerns

---
title: The Fremont Frontier
subtitle: Living at the Margins of Maize Farming
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Peter M. Yaworsky
    affiliations:
      - id: aarhus-culture
        address: |
          Department of Culture and Heritage Studies, 
          School of Culture and Society, 
          Aarhus University, 
          Moesgård Allé 20, building 4216, 
          8270 Højbjerg, Denmark   
      - id: aarhus-biodiversity
        address: |
          Center for Biodiversity Dynamics in a Changing World, 
          Department of Biology, 
          Aarhus University, 
          Ny Munkegade 114-116, 
          8000 Aarhus C, Denmark
  - name: Weston McCool
    affiliations:
      - id: utah-anthropology
        address: | 
          University of Utah Department of Anthropology, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
  - name: Jerry D. Spangler
    affiliations:
      - ref: utah-anthropology
  - name: Simon Brewer
    affiliations:
      - id: utah-geography
        address: |
          University of Utah Department of Geography, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
      - id: gcsc
        address: |
          University of Utah Global Change and Sustainability Center, 
          115 South 1460 East, RM 234 FASB, 
          Salt Lake City, UT 84112
  - name: Brian F. Codding
    affiliations:
      - ref: utah-anthropology
      - ref: gcsc
abstract: |
  The Fremont provide an important case study to examine the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear. Using a tree-ring and simulation-based reconstruction of average annual precipitation and maize growing degree days, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin. The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
keywords:
  - Ideal Free Distribution
  - Great Basin
  - Paleoclimate Reconstruction
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Quaternary International
      formatting: preprint
      model: 3p
execute: 
  echo: false
  error: false
  warning: false
---

@@

Archaeological populations of subsistence maize farmers known collectively as "the Fremont" lived at the far northern periphery of maize farming in western North America, in an area encompassing much of the modern state of Utah north of the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989; @madsen1998]. That the Fremont managed to thrive for as long as they did in this part of the world poses an interesting question, for conditions there are extremely harsh and inhospitable for maize farming. Archaeologists interested in that question have typically focused on strategies directly related to farming and subsistence, like irrigation methods, trade, and seasonal high elevation hunting [e.g., @barlow2008; @boomgarden2019; @janetski2002; @hart2021; @madsen1998; @metcalfe1985; @patterson2010; @spangler1993; @morgan2012]. Here, however, we come at the problem from a slightly different angle. We seek to understand not so much *what* they did to make farming effective, but *where* they chose to do it, and why they chose those places over others [e.g., @bocinsky2014; @thomson2020]. Those are not entirely separate questions, of course, as they trade-off each other, but we think a geographic analysis can actually help illuminate some of the costs and benefits of the other strategies adopted by the Fremont. A geographic or spatial model will also help us to investigate the ways that climate change might have structured those costs and benefits, and by extension the settlement decisions of Fremont farmers, which we consider a long-term record of human adaptation to climate change.

As a test case, we focus on those Fremont living in the eastern Great Basin of western Utah, which we refer to as the western Fremont (see @fig-overview). The extreme desert conditions that the western Fremont adapted to provide a useful backdrop for exploring Fremont settlement, for variation in that environment exhibits abrupt and quite dramatic changes over short distances of space and time, thus heightening the differences between areas the western Fremont did and did not occupy. We explore those differences from a socio-ecological perspective, focusing on dynamic interactions between competing individuals on the one hand and individuals and their environment on the other [@bird2006; @kennett2006a]. This dynamic is captured nicely by the Ideal Free Distribution model from Behavioral Ecology, as it represents individual decisions about where to live as a choice between habitats whose benefits to the individual may be constrained by what other individuals are doing. The technical term for these benefits is 'suitability', which the IFD defines as some function of both the habitat's intrinsic or pristine environmental quality and its current population size [@fretwell1969; @codding2015; @winterhalder2010]. The fundamental assumption underlying the IFD is that individuals will behave optimally, that they will choose to settle the habitat with the highest suitability first, meaning the one that offers them the greatest ratio of benefits to costs relative to the available alternatives.

Assuming that western Fremont individuals were adept at responding to variation in their local ecology, in particular, that they optimized their choice of habitats, we can use an empirical or inductive species distribution model of Fremont sites to reconstruct the ecological niche for maize farming [@vernon2021; @yaworsky2020; @elith2009]. This is sometimes referred to as ecological niche modeling. The basic idea here is that we can look at the distribution of Fremont sites as a reliable indicator of the sorts of available environmental conditions that would best promote maize farming in an arid landscape. In effect, we are helping ourselves to a little reverse engineering of the archaeological record [@dennett1995]. We are assuming that whatever strategy - whatever habitat - the Fremont chose was the best alternative available to them, not in any absolute sense, of course, just the best given the constraints and trade-offs Fremont farmers faced in those environments [@jochim2022]. In other words, we are treating their settlement decisions themselves as maize farming adaptations.

![
On the left is an overview map showing the project area with modern satellite imagery. Dark green colors represent high elevation, high precipitation areas. The large beige area in the northwest is the Bonneville Basin. What remains of the Great Salt Lake can be seen just west of Salt Lake City. For visualization purposes, the map on the right shows the relative density of archaeological sites (see the section on site data below for details on how we calculate relative density). Please note that although the map appears to show many empty watersheds, most do in fact contain archaeological sites, just at extremely low densities.
](../figures/overview.png){#fig-overview fig-align="center" width="6in"}

@@

Archaeologists have nominated a number of artifact types and artifact properties as candidate traits to distinguish the Fremont from their neighbors in space and time, including metates with secondary grinding surfaces, large semi-subterranean pithouses, various clay figurines, trapezoidal rock art, one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and elongated corner-notched arrow points [@madsen1998]. None of these, however, nor any combination of them, has yet proven up to the task of bringing these diverse people under a single, all-encompassing definition [@madsen1998]. Nevertheless, the Fremont do stand out, especially at their population apex roughly 1000 years BP.

Around this time, the Fremont likely reached their greatest geographic extent, inhabiting an area that encompassed most of the modern state of Utah [@janetski2011], with varying levels of support for occupations in Idaho [@dean1992prehistoric], Nevada [@hockett1998], Colorado [@baker1999], and Wyoming [@hakiel1987archery; @smith1992fremont]. These modern political boundaries roughly overlap with the eastern Great Basin and Colorado Plateau physiographic provinces.

![
Linear responses of average water-year precipitation (mm), average growing-season maize gdd (C), and cost-distance to streams (hours) to changes in elevation (m). See methods for how these values were calculated and results for more details.
](../figures/elevation-everything.png){#fig-elevation fig-align="center" width="4in"}

@@

These are the environments into which the Fremont first emerged in western Utah. Models for describing their origins, while diverse, tend to fall into the familiar categories of migration and diffusion. Where models of the former migration variety insist that Ancestral Puebloan farmers migrated northward from the US Southwest, bringing with them drought-tolerant maize varieties and maize farming [@madsen1975; @kidder1924], models of the latter diffusion variety suggest that the Fremont first emerged within these areas as local Archaic populations transitioned away from mobile foraging, adopting semi-sedentary agricultural practices from the southwest in a piecemeal fashion over several hundred years [@jennings1978; @winter1974]. While archaeologists have marshaled an impressive array of arguments and counter-arguments for these two classes of models, a general consensus appears to be coalescing around the idea that the truth lies somewhere in between, that the interaction between a nascent Archaic population existing at extremely low population levels and a burgeoning community of migrant farmers together produced the complex we now call the Fremont [@patterson2015; @simms1986; @spangler1993; @spangler2000; @spangler2013].

Models that try to explain *why* Fremont maize farmers emerged in the far northern periphery of the US Southwest tend to agree - more or less - that at some point farming became a more economical alternative to foraging in the region. Where they differ, it is in whether things got better for farming starting roughly 2000 years BP [@benson2011; @coltrain2002; @matson1988] or worse for foraging [@barlow2002; @broughton2010; @cannon2000; @cannon2001]. In support of the *farming-more-better* model, archaeologists draw on evidence for a suite of favorable climatic conditions that came together around 1600 years BP, including warmer temperatures and increased precipitation (both winter and summer), as well as the expansion of grasslands [@grayson2006; @hemphill1995; @rhode2000]. The greatest expansion of the Fremont, in fact, occurred around 1000 years BP, around the time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures that would have increased the intensity of summer monsoons in the southwest, pushing them further north into the Great Basin and onto the Colorado Plateau [@grayson2006], a pattern reflected in numerous tree ring records [e.g., @graybill1990; @leavitt1994; @stine1990; @salzer2014], though some complications do exist for this suggested pattern [@hart2021].

In response to these sorts of arguments, those who favor the *foraging-more-worse* model [@hart2021; @barlow2002] argue that conditions promoting maize farming would also tend to make foraging a more productive strategy. Plus, farming is an extremely costly endeavor, especially when compared to foraging, including lots of upfront investment of time and energy, so improving conditions favorable to maize agriculture are not by themselves sufficient to explain its origins. To fill that lacuna, archaeologists point to accumulating evidence suggesting that the Fremont transition to agriculture accompanied periods of sustained population growth [@barlow2008; @codding2021; @spangler2013] that likely resulted in the depressed availability of wild resources [@barlow2002; @janetski1997; @simms1986]. So, demographic pressure reduced the efficiency of foraging strategies, thus making maize agriculture the optimal strategy. And that, in turn, led the Fremont to adopt farming as their primary mode of subsistence.

Once they entered the farming niche, the Fremont quickly developed several adaptations for coping with the marginal and stochastic conditions in the eastern Great Basin and the Colorado Plateau, the most important almost certainly being irrigation (for more on this, see the discussion). On the Colorado Plateau, the eastern Fremont settled deep within the steep canyons and washes that flanked the tributaries of major rivers like the Green and Yampa [@yaworsky2021; see also, Yaworsky this issue], in places like Nine Mile Canyon [@spangler1993] and Range Creek Canyon [@hart2021]. In the eastern Great Basin, the western Fremont chose to settle on open, alluvial fans and stream terraces along the lower slopes of most ranges [@janetski2000; @madsen1998; @simms2008], at places like Nephi Mounds [@sharrock1967], Pharo Village [@marwitt1968], Median Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock Village [@aikens1967].

Given the extreme aridity of the west, they were still susceptible to drought, of course, even in those more suitable areas, with a single short growing season making agricultural returns highly variable. To cope with these uncertainties, Fremont farmers may have adopted additional risk-mitigation strategies like spreading farm plots among different micro-environments, a technique otherwise known as plot diversification [@spangler1993; @spangler2013; @patterson2015]. The Fremont also relied heavily on crop storage both to provide a surplus during the winter season and to make-up for poor yields [@spangler2019].

## The Ideal Free Distribution

In population ecology, settlement decisions are represented as a choice between habitats, where each habitat $i$ has some net biological value or benefit, known as its suitability ($S_i$). Typically, though not always, this value exhibits negative density dependence, declining as a function of population size ($N_i$), perhaps owing to increased competition or resource depression. When $N_i$ is very small, the suitability of a habitat approaches its pristine or intrinsic quality ($Q_i$), or its gross value independent of all settlement costs [@greene2001]. Together, these parameters suggest a very simple model of habitat suitability:

$$
S_i = Q_i - f(N_i)
$$

This is known as the Ideal Free Distribution (IFD) model [@fretwell1969]. The IFD model assumes that individuals will choose habitats that maximize $S_i$, settling the most suitable habitat first then shifting to the next most suitable habitat when those become equal, and thereafter infilling each at an equal rate. The aggregate effect is a population distributed at equilibrium, meaning suitability is equal for all habitats. When conditions are ideal (individuals have perfect knowledge) and free (individuals face no additional costs to settlement), this distribution will also satisfy the "input matching" rule, with populations occurring at larger numbers in habitats with greater intrinsic value [@parker1978].

Two circumstances provide important exceptions to this simple model. The first concerns economies of scale, or situations in which $S_i$ exhibits positive density dependence. In ecology, these are known as Allee effects, and they are assumed to occur at low density [@allee1949; @fretwell1969]. The second concerns additional settlement costs, or costs associated with moving from one habitat to another. This is a violation of the free element of the IFD. When it occurs, the result is an ideal despotic distribution (IDD) [@fretwell1969]. This has the consequence of reducing the total population in the more suitable habitat at equilibrium.

Many social and ecological variables may be relevant to defining a general measure of $S_i$, but here we are focusing on ecological determinants of intrinsic habitat quality, $Q_i$, that are specific to maize farmers. Our reasoning here is fairly straightforward. If we assume that the distribution of Fremont populations across habitats is at equilibrium, then we can combine this concept of maize-specific suitability with the input matching rule to reconstruct the ranking of habitats on the basis of ecologically-relevant parameters. That is, it follows from their being at equilibrium that any observed differences in population size must be the result of differences in $Q_i$, that $Q_i \approx f(N_i)$ at equilibrium. So, if we have some proxy for $N_i$, like the distribution of Fremont sites across habitats, we can model that value as a function of ecological covariates $X_i$, then use the resulting model to reconstruct $Q_i \approx f(X_i)$, giving us the maize niche for Fremont farmers, both in geographic and ecological space.

@@

Because we are focusing on the western Fremont, we constrain the current project area to the eastern Great Basin in western Utah, as shown in @fig-overview, where the Great Basin is defined in the hydrologic sense using watershed boundaries [@grayson2011]. On the far eastern edge of this region, a series of north-south ranges, including the Wasatch, Pahvant, and Tushar Mountains, together stand as the primary physiographic boundaries separating the Great Basin from the Colorado Plateau. On its far western edge, a series of north-south ranges straddle the Nevada state line or lie just inside it, including the Snake Range where Great Basin National Park is located. Its northern periphery is the Great Salt Lake and the West Desert, its southern boundary the Pine Valley Mountains that separate the Great Basin from the Virgin River watershed, just north of St. George, UT. This is a region approximately 107,000 km$^2$ in area, encompassing virtually all of the Bonneville Basin in western Utah.

>>>>

---
title: The Fremont Frontier
subtitle: Living at the Margins of Maize Farming
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Peter M. Yaworsky
    affiliations:
      - id: aarhus-culture
        address: |
          Department of Culture and Heritage Studies, 
          School of Culture and Society, 
          Aarhus University, 
          Moesgård Allé 20, building 4216, 
          8270 Højbjerg, Denmark   
      - id: aarhus-biodiversity
        address: |
          Center for Biodiversity Dynamics in a Changing World, 
          Department of Biology, 
          Aarhus University, 
          Ny Munkegade 114-116, 
          8000 Aarhus C, Denmark
  - name: Weston C. McCool
    affiliations:
      - id: utah-anthropology
        address: | 
          University of Utah Department of Anthropology, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
  - name: Jerry D. Spangler
    affiliations:
      - ref: utah-anthropology
  - name: Simon Brewer
    affiliations:
      - id: utah-geography
        address: |
          University of Utah Department of Geography, 
          260 S. Central Campus Drive, RM 4625, 
          Salt Lake City, UT 84112
      - id: gcsc
        address: |
          University of Utah Global Change and Sustainability Center, 
          115 South 1460 East, RM 234 FASB, 
          Salt Lake City, UT 84112
  - name: Brian F. Codding
    affiliations:
      - ref: utah-anthropology
      - ref: gcsc
abstract: |
  The Fremont provide an important case study for examining the resilience of ancient farmers to climatic downturns, for they lived at the far northern margin of intensive maize agriculture in the American West, where the constraints on maize production are made abundantly clear. Using a tree-ring and simulation-based reconstruction of average annual precipitation and temperature, along with cost-distance to perennial streams, we model spatial variability in Fremont site density in the eastern Great Basin. The results of our analysis have implications for defining the ecological envelope in which farming is a viable strategy across this arid region and can be used to predict where and why maize farming strategies might evolve and eventually collapse as climate changes over time.
keywords:
  - Ideal Free Distribution
  - Great Basin
  - Paleoclimate Reconstruction
corresponding_author: Kenneth.Vernon@colorado.edu
linenumbers: false
bibliography: bibliography.bib
format: 
  simple-pdf: default
  docx: default
execute: 
  echo: false
  error: false
  warning: false
---

@@

Archaeological populations of subsistence maize farmers known collectively as
"the Fremont" lived at the far northern periphery of maize farming in western
North America, in an area encompassing much of the modern state of Utah north of
the Virgin and Colorado Rivers, from roughly 2000 to 700 years BP [@madsen1989;
@madsen_simms1998]. Conditions in this area are extremely harsh and inhospitable
to maize farming, so the fact that the Fremont not only pulled it off but also
thrived while doing so poses an interesting puzzle. How did they do it?
Archaeologists interested in this question have typically focused on strategies
directly related to farming and subsistence, like irrigation methods, trade, and
seasonal high elevation hunting [e.g., @barlow_etal2008; @boomgarden_etal2019;
@janetski2002; @hart_etal2021; @madsen_simms1998; @metcalfe_larrabee1985;
@patterson_flanigan2010; @spangler1993; @morgan_etal2012]. Here, however, we
come at the problem from a slightly different angle. We seek to understand not
so much *what* they did to make farming effective, but *where* they chose to do
it, and why they chose those places over others [e.g., @bocinsky_kohler2014;
@thomson_macdonald2020]. Those are not entirely separate questions, of course,
as they trade-off each other, but we think a geographic analysis can actually
help illuminate some of the costs and benefits of the other strategies adopted
by the Fremont. A geographic approach will also help us to investigate the ways
that climate change might have structured those costs and benefits, and by
extension the settlement decisions of Fremont farmers.

As a test case, we focus on those Fremont living in the eastern Great Basin of
western Utah, which we refer to as the western Fremont (see @fig-overview). The
extreme desert conditions that the western Fremont adapted to provide a useful
backdrop for exploring Fremont settlement, for variation in that environment
exhibits abrupt and quite dramatic changes over short distances of space and
time, thus heightening the differences between areas the western Fremont did and
did not occupy. We explore those differences from a socio-ecological
perspective, focusing on dynamic interactions between competing individuals on
the one hand and individuals and their environment on the other
[@bird_oconnell2006; @kennett_etal2006]. This dynamic is captured nicely by the
Ideal Free Distribution model from Behavioral Ecology, as it represents
individual decisions about where to live as a choice between habitats whose
benefits to the individual may be constrained by what other individuals are
doing. The technical term for these benefits is 'suitability', which the IFD
defines as some function of both the habitat's intrinsic or pristine
environmental quality and its current population size [@fretwell_lucas1969;
@codding_bird2015; @winterhalder_etal2010].

The fundamental assumption underlying the IFD is that individuals will behave
optimally, that they will choose to settle the habitat with the highest
suitability first, meaning the one that offers them the greatest ratio of
benefits to costs relative to the available alternatives. Of course, this comes
with the usual caveats. The claim is not that individuals are infallible, that
they will always do exactly the right thing, but that their choices will tend to
approximate the best strategy given their constraints and trade-offs.

Assuming the western Fremont optimized their agricultural practices, we can use
an empirical or inductive species distribution model [SDM, @elith_etal2006;
@elith_leathwick2009] of Fremont sites to estimate relative differences in maize
suitability across specific environmental gradients and geographic locations
[@jochim2022; @vernon_etal2022a; @yaworsky_etal2020; @elith_leathwick2009]. In
ecology, you will sometimes hear this approach referred to as "ecological niche
modeling" [e.g., @feng_etal2019; @sillero_barbosa2021; @sillero_etal2021] or
even more explicitly as "habitat suitability modeling" [e.g., @bowden_etal2021;
@rowden_etal2017; @dekort_etal2020]. The basic idea here is that we can look at
the distribution of Fremont sites as a reliable indicator of the sorts of
available environmental conditions that would best promote maize farming in an
arid landscape. To put that in more blatantly economic terms, we are assuming
that their environmental preferences are shaped primarily by the constraints of
maize farming and revealed in their residential choice behavior. This allows us
to build a tentative model of the ecological niche for maize farming and to make
some defeasible inferences about its limiting conditions.

::: {#fig-overview fig-align="center"}
![](../figures/overview.jpg){width="5in"}

On the left is an overview map showing the project area with modern satellite
imagery, so the green areas indicate plant growth, the beige areas, empty
desert. The green areas are also the mountainous areas. Thick orange lines are
interstate highways. Thin orange lines are major state highways. For
visualization purposes, the map on the right shows the log-transformed density
of feature-weighted archaeological sites. If no sites have been recorded in a
watershed, log(1e-5) is shown.
:::

@@

Archaeologists have nominated a number of artifact types and artifact properties
as candidate traits to distinguish the Fremont from their neighbors in space and
time, including metates with secondary grinding surfaces, large
semi-subterranean pithouses, various clay figurines, trapezoidal rock art,
one-rod-and-bundle basketry, painted white and corrugated gray ceramics, and
elongated corner-notched arrow points [@madsen_simms1998]. None of these,
however, nor any combination of them, has yet proven up to the task of bringing
these diverse people under a single, all-encompassing definition
[@madsen_simms1998]. Nevertheless, the Fremont do stand out, especially at their
population apex roughly 1000 years BP.

Around this time, the Fremont likely reached their greatest geographic extent,
inhabiting an area that encompassed most of the modern state of Utah
[@janetski_etal2011], with varying levels of support for Fremont foragers in
Idaho [@dean1992prehistoric] and Wyoming [@hakiel1987archery; @smith1992fremont]
and Fremont farmers in Nevada [@hockett1998; @cole2012a] and Colorado
[@baker1999]. These modern political boundaries roughly overlap with the eastern
Great Basin and Colorado Plateau physiographic provinces.

::: {#fig-elevation fig-align="center"}
![](../figures/elevation-everything.jpg){width="4in"}

Linear responses of the centered and scaled average water-year precipitation
(mm), average growing-season maize gdd (C), and cost-distance to streams (hours)
to changes in elevation (m). See methods for how these values were calculated
and results for more details.
:::

@@

These are the environments into which the Fremont first emerged in western Utah.
Models for describing their origins, while diverse, tend to fall into the
familiar categories of migration and diffusion. Where models of the former
migration variety insist that Ancestral Puebloan farmers migrated northward from
the US Southwest, bringing with them drought-tolerant maize varieties and maize
farming [@madsen_berry1975; @kidder1924], models of the latter diffusion variety
suggest that the Fremont first emerged within these areas as local Archaic
populations transitioned away from mobile foraging, adopting semi-sedentary
agricultural practices from the Southwest in a piecemeal fashion over several
hundred years [@jennings1978; @winter_wylie1974]. While archaeologists have
marshaled an impressive array of arguments and counter-arguments for these two
classes of models, a general consensus appears to be coalescing around the idea
that the truth lies somewhere in between, that the interaction between a
lingering Archaic population existing at extremely low population levels and a
burgeoning community of migrant Southwest farmers together produced the complex
we now call the Fremont [@patterson2015; @spangler1993; @spangler2000;
@spangler2013; @searcy_talbot2015; @simms2008].

Models that try to explain *why* Fremont maize farmers emerged in the far
northern periphery of the US Southwest tend to agree - more or less - that at
some point farming became a more economical alternative to foraging in the
region. Where they differ, it is in whether things got better for farming
starting roughly 2000 years BP [@benson2011; @coltrain_leavitt2002;
@matson_etal1988] or worse for foraging [@barlow2002; @broughton_etal2010;
@cannon2000; @cannon2001]. In support of the *farming-got-better* model,
archaeologists draw on evidence for a suite of favorable climatic conditions
that came together around 1600 years BP, including warmer temperatures and
increased precipitation (both winter and summer), as well as the expansion of
grasslands [@grayson2006; @hemphill_wigand1995; @madsen2000]. The greatest
expansion of the Fremont, in fact, occurred around 1000 years BP, around the
time the Medieval Climate Anomaly (MCA) introduced warmer summer temperatures
that would have increased the intensity of summer monsoons in the Southwest,
pushing them further north into the Great Basin and onto the Colorado Plateau
[@grayson2006], a pattern reflected in numerous tree ring records [e.g.,
@graybill1990; @leavitt1994; @stine1990; @salzer_etal2014], though some
complications do exist for this suggested pattern [@hart_etal2021].

In response to these sorts of arguments, those who favor the
*foraging-got-worse* model [@hart_etal2021; @barlow2002] argue that conditions
promoting maize farming would also tend to make foraging a more productive
strategy. Plus, farming is an extremely costly endeavor, especially when
compared to foraging, including lots of upfront investment of time and energy,
so improving conditions favorable to maize agriculture are not by themselves
sufficient to explain its origins. To fill that lacuna, archaeologists point to
accumulating evidence suggesting that the Fremont transition to agriculture
accompanied periods of sustained population growth [@barlow_etal2008;
@codding_etal2022; @spangler2013] that likely resulted in the depressed
availability of wild resources [@barlow2002; @janetski1997; @simms1986]. So,
demographic pressure reduced the efficiency of foraging strategies, thus making
maize agriculture a more profitable alternative. And that, in turn, led the
Fremont to adopt farming as their primary mode of subsistence.

Once they entered the farming niche, the Fremont quickly developed several
adaptations for coping with the marginal and stochastic conditions in the
eastern Great Basin and the Colorado Plateau, the most important almost
certainly being irrigation (for more on this, see the discussion). On the
Colorado Plateau, the eastern Fremont settled deep within the steep canyons and
washes that flanked the tributaries of major rivers like the Green and Yampa
[@yaworsky2021], in places like Nine Mile Canyon [@spangler1993] and Range Creek
Canyon [@hart_etal2021]. In the eastern Great Basin, the western Fremont chose
to settle on open, alluvial fans and stream terraces along the lower slopes of
most ranges [@janetski_talbot2000; @madsen_simms1998; @simms2008], at places
like Nephi Mounds [@sharrock_marwitt1967], Pharo Village [@marwitt1968], Median
Village [@marwitt1970], Evans Mound [@berry1972; @dodd1982], and Snake Rock
Village [@aikens1967].

Given the extreme aridity of the west, they were still susceptible to drought,
of course, even in those more suitable areas, with a single short growing season
making agricultural returns highly variable. To cope with these uncertainties,
Fremont farmers may have adopted additional risk-mitigation strategies like
spreading farm plots among different micro-environments, a technique otherwise
known as plot diversification [@spangler1993; @spangler2013; @patterson2015].
The Fremont also relied heavily on crop storage both to provide a surplus during
the winter season and to make-up for poor yields [@spangler_etal2019].

@@

Because we are focusing on the western Fremont, we constrain the current project
area to the eastern Great Basin in western Utah, as shown in @fig-overview,
where the Great Basin is defined in the hydrologic sense using watershed
boundaries [@grayson2011]. On the far eastern edge of this region, a series of
north-south ranges, including the Wasatch Mountains in the north and the Sevier
Plateau in the south, together stand as the primary physiographic boundaries
separating the majority of the Great Basin from the Colorado Plateau. On its far
western edge, a series of north-south ranges straddle the Nevada state line or
lie just inside it, including the Snake Range where Great Basin National Park is
located. Its northern periphery includes the Great Salt Lake (GSL) and the West
Desert, its southern boundary the Bull and Pine Valley Mountains that separate
the Great Basin from the Virgin River watershed just north of St. George, UT.
This is a region approximately 107,000 km$^2$ in area, encompassing virtually
all of the Bonneville Basin in western Utah.
